Abstractions for Multi-Sorted Substitutions

### 2023-09-16 paper/paper.tex
Update.

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, operations on terms have to map
  well-typed terms to well-typed terms proving that they preserve typing.
  For example, in an intrinsic setting, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  When working without library support, this often has the benefit of
  more concise code, as shared structure between definitions and
  preservation lemmas does not need to be repeated.
  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.

>>>>

  With \emph{intrinsic typing}, the definitions of syntax and typing
  are merged into a single definition of well-typed terms.
  Consequently, operations on terms have to map
  well-typed terms to well-typed terms proving that they preserve typing.
  For example, the definition of substitution
  also acts as a proof that substitution preserves typing, and the
  definition of a small-step semantics also acts as a proof for subject
  reduction.
  When working without library support, intrinsic typing often has the
  benefit of more concise code, as shared structure between definitions
  and preservation lemmas does not need to be repeated.

  With \emph{extrinsic typing}, the definitions of syntax and typing
  are kept separately: the syntax is defined first, and then typing is
  defined as a relation. Consequently, operations on terms and
  their type preservation lemmas are also separate entities.

  However, when working with library support, this is less of a benefit.
  Most operations, like substitution, have a generic definition and
  metatheory across languages, which can be derived automatically via
  reflection or generic programming techniques.
  Only by adding a typing-relation

  However, intrinsic typing also has three major drawbacks:
  \begin{enumerate}
  \item
    It is more difficult to judge whether a specification is correct,
    because there is no clear separation between specification and proofs.
    To understand the specification one has to additionally
    understand the proofs, which can introduce significant cognitive overhead.
    % For example specifying a small step semantics is directly intertwined
    % with the proof of subject reduction.
  \item
    It is less uniform for object languages with multiple variable
    sorts, e.g. type- and expression-variables. In such a scenario,
    types are usually indexed by kinds and expressions by kinds and types.
    This requires not only a type-in-type and expression-in-expression
    substitution, but also a type-in-expression substitution, which causes
    interactions between different types of substitutions introducing
    additional complexity.
    The duplication increases even more if one additionally introduces
    kind-variables. Then a total of 6 different substitutions and
    their interactions is required.
    Unifying these substitutions to reduce repetition is difficult
    with intrinsic typing due to the interdependence of definitions.
  \item
    It is less suitable for automation, because type preservation is baked
    into the syntax definition and type preservation can become arbitrary complex
    depending on the typing relation. This can prevent substitition
    definition and lemmas from being derived, where otherwise it would be possible.
  \end{enumerate}
  With extrinsic typing the major drawback is the duplication between
  definitions and type preservation lemmas, which we found more
  suitable to be eliminated by automation.
